openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding

### 2020-07-06 paper/policar2020-opentsne.tex
Revise manuscript

%We implement t-SNE~\cite{maaten2008visualizing}. We include a set of recommendations and tricks from Kobak and Berens~\cite{kobak2019art} as well as incerasing the learning rate as recommended by Belkina \textit{et al.}~\cite{belkina2019automated}. Also support using variable degrees of freedom by Kobak \textit{et al.}~\cite{kobak2019heavy}.
%
%Two approximation schemes, one based on Barnes-Hut~\cite{van2014accelerating} and one on the interpolation-based approach~\cite{linderman2019fast}.
%
%UMAP~\cite{2018arXivUMAP}, whether or not it better preserves global structure or not is in question~\cite{kobak2019umap}.

%% TODO: START FROM FIRST PAPER

The abundance of high-dimensional data sets in molecular biology calls for
techniques for dimensionality reduction, and in particular for methods that can
help in the construction of data visualizations. Popular approaches for
dimensionality reduction include principal component analysis, multidimensional
scaling, and uniform manifold approximation and projections~\cite{2018arXivUMAP}. Among
these, t-distributed stochastic neighbor embedding
(t-SNE)~\cite{maaten2008visualizing} lately received much attention as it can
address high volumes of data and reveal meaningful clustering structure. Most
of the recent reports on single-cell gene expression data start with an
overview of the cell landscape, where t-SNE embeds high-dimensional expression
profiles into a two-dimensional space~\cite{macosko2015highly,cao2019single,tasic2018shared}.
Fig.~\ref{fig:tsne}.a presents an example of one such embedding.

Despite its utility, t-SNE has been criticized for poor scalability
when addressing large data sets, lack of global organization \em t-SNE
focuses on local clusters that are arbitrarily scattered in the
low-dimensional space \em and absence of theoretically-founded implementations to
map new data into existing embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most of these shortcomings, have
recently been addressed.
\cite{fi_tsne} sped-up the method through an interpolation-based
approximation, reducing the time complexity to be merely linear dependent on the number of samples. \cite{art_of_using_tsne} proposed
several techniques to improve global positioning, including estimating similarities with a mixture of
Gaussian kernels. While no current popular
software library supports mapping of new data into reference embedding,
\cite{parametric_tsne} proposed a related approach
using neural networks.

%% TODO: END FROM FIRST PAPER

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{macosko2015}
  \caption{\label{fig:macosko}}
\end{figure*}

@@

We introduce openTSNE, a comprehensive Python library that
implements t-SNE and all its recently proposed extensions. The library is
compatible with the Python data science ecosystem (e.g., {\tt numpy}, {\tt
sklearn}, {\tt scanpy}). Its modular design encourages extendibility and
experimentation with various settings and changes in the analysis pipeline.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{tasic2018}
  \caption{\label{fig:tasic}We use openTSNE to create four different visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data set containing 21,874 single-cells originating from the mouse neocortex. Cluster annotations and colors are taken from the original publication. Warm colors correspond to excitatory neurons, cool colors correspond to inhibitory neurons, and gray/brown colors correspond to non-neuronal cells. The standard t-SNE \textbf{(a)} emphasizes local structure, while increasing perplexity \textbf{(b)} results in a more meaningful layout of the clusters. We can also combine the two perplexities by using a multiscale kernel affinity model \textbf{(c)} and obtain a good trade-off between global and local structure. Alternatively, we can inspect more fine-grained structure and reveal smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}

The following code is used to generate Fig.~\ref{fig:tasic}
\begin{minted}[fontsize=\footnotesize]{python}
# Perplexity 30 (standard t-SNE)
perp_30 = openTSNE.TSNE(
    perplexity=30, metric="cosine"
).fit(X)

# Perplexity 500 (emphasise global structure)
perp_500 = openTSNE.TSNE(
    perplexity=500, metric="cosine"
).fit(X)

# Multiscale kernel with perplexities 30,500
affinities = openTSNE.affinity.Multiscale(
    X, perplexities=[30, 500], metric="cosine"
)
init = openTSNE.initialization.pca(X)
ms = openTSNE.TSNEEmbedding(init, affinities)
ms = ms.optimize(n_iter=250, exaggeration=12)
ms = ms.optimize(n_iter=250, exaggeration=1)

# Expose finer structure using smaller dof
perp_30 = openTSNE.TSNE(
    perplexity=30, metric="cosine", dof=0.6
).fit(X)
\end{minted}

@@

The following code is used to generate Fig.~\ref{fig:cao}
\begin{minted}[fontsize=\footnotesize]{python}
indices = np.random.permutation(list(range(x.shape[0])))
reverse = np.argsort(indices)
# Split the data into a sample/remainder sets
x_sample, x_rest = x[indices[:25000]], x[indices[25000:]]
y_sample, y_rest = y[indices[:25000]], y[indices[25000:]]
# Create an initial embedding using a subset of the data
sample_embedding = openTSNE.TSNE(
    initialization="spectral",
    perplexity=500,
    metric="cosine",
).fit(x_sample)
# Position the remainder of the points to their nearest
# neighbor in the sample embedding
rest_init = sample_embedding.prepare_partial(x_rest, k=1)
init_full = np.vstack(
    [sample_embedding, rest_init]
)[reverse]
init_full = openTSNE.initialization.rescale(init_full)
# Calculate the affinity model using all the data points
affinities = openTSNE.affinity.PerplexityBasedNN(
    x, perplexity=30, metric="cosine"
)

embedding = openTSNE.TSNEEmbedding(
    init_full, affinities
%)
exag12 = embedding.optimize(n_iter=500, exaggeration=12)
exag4 = exag12.optimize(n_iter=500, exaggeration=4)
exag2 = exag4.optimize(n_iter=500, exaggeration=2)
exag1 = exag2.optimize(n_iter=500, exaggeration=1)
\end{minted}

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{transform_hochgerner}
  \caption{\label{fig:transform}openTSNE enables the user to embed new samples into an existing, reference embedding. We construct a t-SNE embedding for the data from Hochgerner \textit{et al.}~\cite{hochgerner2018conserved} containing 24,185 developing, single cells from the mouse hippocampus. The data contains different types of neurons, supporting glia, and other vascular cells (upper left). Cell annotations are taken from the original publications. The colors correspond to Fig.~\ref{fig:tasic}. We then embed new, hippocampal cells collected by Harris \textit{et al.}~\cite{harris2018classes} using the previous embedding as a reference. In their study, Harris \textit{et al.} collected 6,971 single-cells and focused on identifying different types of inhibitory neurons. However, almost half of the collected cells are not neurons and were left uncharacterized. Inspecting where these cells are embedded in the reference embedding (bottom left) reveals that in addition to inhibitory neurons, the data contains a number of supporting glia as well as a small population of endothelial cells. We can verify the accuracy of our approach by inspecting marker genes for the major cell types in the reference (top row) and embedded samples (bottom row).}
\end{figure*}

The following code is used to generate Fig.~\ref{fig:transform}
\begin{minted}[fontsize=\footnotesize]{python}
# Assuming we have prepared a reference embedding and
# assuming we have aligned new_X to the reference X matrix
new_embedding = reference_embedding.transform(new_X)
\end{minted}

We want to mention that this works for mitigating batch effects in our earlier work~\cite{policar2019embedding}.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{benchmarks}
  \caption{\label{fig:benchmarks}We benchmark openTSNE against three popular open-source implementations including scikit-learn~\cite{pedregosa2011scikit}, MulticoreTSNE~\cite{Ulyanov2016}, and FIt-SNE~\cite{linderman2019fast}. Benchmarks were run on a consumer-grade Intel Core i7-7700HQ processor found in laptop computers, and on a server-grade Intel Xeon E5-2650. To generate benchmark data sets of different sizes, we subsampled data from the 10X Genomics 1.3 million mouse brain data set five times, resulting in five different data sets for each size. In total, each implementation was run on 30 different data sets. Notice that openTSNE scales similarly to FIt-SNE, as they both use the same interpolation-based approximation scheme, while scikit-learn and MulticoreTSNE utilize the Barnes-Hut approximation.}
\end{figure*}

\section*{Discussion}
Text for this section \ldots

@@

% approach name in bold, reference, description of the approach and why is
% it useful. focus on theory and math.

\subsection*{t-SNE}
%% TODO: Reword this entirely

Local, non-linear dimensionality reduction by t-SNE is performed as follows.
Given a multi-dimensional data set $\mathbf{X} = \left \{ \mathbf{x}_1,
\mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in \mathbb{R}^D$ where $N$ is the
number of data points in the reference data set, t-SNE aims to find a low
dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1, \mathbf{y}_2, \dots,
\mathbf{y}_N \right \} \in \mathbb{R}^d$ where $d \ll D$, such that if points
$\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the multi-dimensional space,
their corresponding embeddings $\mathbf{y}_i$ and $\mathbf{y}_j$ are also
close. Since t-SNE is primarily used as a visualization tool, $d$ is typically
set to two. The similarity between two data points in t-SNE is defined as:

@@

\noindent where $\mathcal{D}$ is a distance measure. This is then symmetrized to

@@

\noindent Different bandwidths $\sigma_i$ enable t-SNE to adapt to the varying
density of the data in the multi-dimensional space.

The similarity between points $\mathbf{y}_i$ and $\mathbf{y}_j$ in the
embedding space is defined using the $t$-distribution with one degree of
freedom

@@

The t-SNE method finds an embedding $\mathbf{Y}$ that minimizes the
Kullback-Leibler (KL) divergence between $\mathbf{P}$ and $\mathbf{Q}$,

@@

It is then easy to show that the corresponding gradient takes on the form
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j || ^2 \right )^{-1}.
\label{eq:tsne_gradient}
\end{equation}

Optimization is performed using batch gradient descent with the delta-bar-delta update rule for increased convergence speed~\cite{jacobs1988increased}. Originally, t-SNE was run for 1000 iterations consisting of two phases. In the first \textit{early exaggeration} phase, the attractive forces between data points is increased by a factor $\rho=12$ so that points can move throughout the embedding uninhibited by repulsive forces. Then the remaining 750 iterations are run with $\rho=1$, which reverts the attractive forces to their original values.

Belkina \textit{et al.} later found that the number of iterations can saftely be decreased to 750 iterations by increasing the learning rate to $\eta=N/12$ instead of the default and somewhat arbitrarily set $\eta=200$~\cite{belkina2019automated}. A welcome side effect of this reduction is also faster runtime.

@@

Van der Maaten observed that since $p_{ij}$ is calculated in the high-dimensional space using a Gaussian kernel, and that the bandwidth is selected such that only a fixed number of closest neighbors (determined by the perplexity parameter) will contribute to the attractive forces of a given point due to the exponential decay of the Gaussian kernel. Therefore it is sufficient to calculate the attractive forces only over a relatively small number of nearest neighbors instead of all $N$ points. By utilizing tree-based nearest-neighbor search methods, the time complexity is thus reduced to $\mathcal{O}(N \log N)$. Linderman \textit{et al.} further realised that, qualitatively, embeddings are visually virtually indistinguishable when using only \textit{approximate} nearest neighbors, further reducing time complexity to $\mathcal{O}(N)$~\cite{linderman2019fast}.

@@

openTSNE is currently the only open-source t-SNE implementation allowing users to add new samples into existing embeddings.

@@

In standard t-SNE $\alpha=1$ so this simplifies to Eqn.~(\ref{eq:cauchy_kernel}).

>>>>

The abundance of high-dimensional data sets in molecular biology calls for
techniques for dimensionality reduction, and in particular for methods that can
help in the construction of data visualizations. Popular approaches for
dimensionality reduction include principal component analysis (PCA), multidimensional
scaling, t-distributed stochastic
neighbor embedding (t-SNE)~\cite{maaten2008visualizing}, and uniform manifold approximation and
projections (UMAP)~\cite{2018arXivUMAP}. Among these, t-SNE lately received much
attention as it can address high volumes of data and reveal meaningful
clustering structure. Most of the recent reports on single-cell gene expression
data start with an overview of the cell landscape, where t-SNE embeds
high-dimensional expression profiles into a two-dimensional
space~\cite{macosko2015highly,cao2019single,tasic2018shared}.
Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b are examples of two such
embeddings.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{macosko2015}
  \caption{\label{fig:macosko}The figure shows the progression of t-SNE embeddings following theoretical advances, which are all included in \opentsne. The data in \textbf{(a)} and \textbf{(b)} represent 44,808 single-cell gene-expression profiles of mouse retinal cells from the study by Macosko \textit{et al.}~\cite{macosko2015highly}. Similarly, the data in \textbf{(c)} contains 27,499 expression profiles from mouse retinal cells from Shekhar \textit{et al.}~\cite{shekhar2016comprehensive}. \textbf{(a)} We construct a t-SNE embedding following the parameter choices from the original publication from Maaten \& Hinton~\cite{maaten2008visualizing}. The visualization shows poor preservation of the global organization of clusters, resulting from random initialization and an affinity model focused on preserving local neighborhoods. \textbf{(b)} shows a modern t-SNE embedding, utilizing the latest theoretical advances and practical recommendations. The embedding is constructed using a multi-scale affinity model, preserving both short-range and long-range interactions between data points, and is initialized such that the global layout is meaningful. Unlike in \textbf{(a)}, the green and blue clusters representing different sub-types of amacrine and bipolar cells are now localized to the same regions of the space, indicating a higher level of similarity than to other cell types. \textbf{(c)} shows how existing t-SNE reference atlases can be used to embed new samples into existing embeddings. The positions of new data points correspond to cell types from the reference atlas.}
\end{figure*}

Despite its utility, t-SNE has often been criticized for poor scalability when
addressing large data sets, lack of global organization -- t-SNE focuses on local
clusters that are arbitrarily scattered in the low-dimensional space -- and the
absence of theoretically-founded implementations to map new data into existing
embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most of these
shortcomings have recently been addressed. Linderman \textit{et
al.} sped-up the method through an interpolation-based
approximation, reducing the time complexity to be merely linear dependent on
the number of samples~\cite{linderman2019fast}. Kobak \& Berens proposed
several techniques to improve global positioning, including estimating
similarities with a mixture of Gaussian kernels~\cite{kobak2019art}. In our previous work, we presented
a principled approach for embedding new samples into existing visualizations~\cite{policar2019embedding}.

@@

We introduce \opentsne, a comprehensive Python library that implements t-SNE and all its recently proposed extensions. Most notable among these are
\begin{enumerate}
\item the implementation of efficient approximation schemes~\cite{van2014accelerating,linderman2019fast} allowing the embedding of millions of data points,
\item the addition of new samples into existing embeddings~\cite{policar2019embedding},
\item better initialization schemes~\cite{kobak2019umap} leading to more globally consistent layouts,
\item variable degrees of freedom~\cite{kobak2019heavy} allowing the inspection of data at different levels of resolution,
\item multiscale similarity kernels~\cite{kobak2019art} which preserves small, well-defined clusters and uncovers global relationships between clusters, and
\item improved defaults for learning rate and the number of iterations~\cite{belkina2019automated} which produce embeddings at a lower computational cost.
\end{enumerate}

The library is fully compatible with the Python data-science ecosystem (e.g., \textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}) providing a familiar and intuitive user interface to new users. Its modular design encourages extensibility and experimentation with various settings and changes in the analysis pipeline.

\opentsne\ is open-source and has become an integral part of the Python data science ecosystem. The package has received over 600 Github stars and has been downloaded hundreds of thousands of times, averaging in thousands of weekly downloads. The source code is publicly available on Github at \url{https://github.com/pavlin-policar/openTSNE}, and can be installed from \textsf{PyPI} and \textsf{conda-forge} -- the two most widely adopted Python package managers.

Accessibility is one of the core design principles of \opentsne. Our aim is to make the latest theoretical advances in t-SNE as accessible as possible to as many users as possible. As such, our API is based on the familiar \textsf{scikit-learn} interface~\cite{sklearn_api}, which enables users a smooth transition from existing Python libraries. For example, Fig.~\ref{fig:tasic} depicts four different t-SNE visualizations generated by \opentsne, each based on recent advances in the t-SNE algorithm. The code used to generate these embeddings is shown below.

\begin{minted}[
fontsize=\footnotesize,
frame=lines,
framesep=2mm,
baselinestretch=1.1
]{python}
import openTSNE
from openTSNE.affinity import Multiscale
# a - standard t-SNE
tsne1 = openTSNE.TSNE(perplexity=30).fit(X)
# b - high perplexity  for globa structure
tsne2 = openTSNE.TSNE(perplexity=500).fit(X)
# c - multiscale kernel for local/global structure
aff = Multiscale(X, perplexities=[30, 500])
tsne3 = openTSNE.TSNE(affinities=aff).fit(X)
# d - decrease dof for higher resolution
tsne4 = openTSNE.TSNE(dof=0.6).fit(X)
\end{minted}

\noindent The code snippet takes as input a \textsf{numpy} array or \textsf{scipy} sparse matrix \texttt{X} and produces four \texttt{TSNEEmbedding} objects. These may subsequently be subject to further optimization under different parameter settings or may be used to embed new samples into the embedding landscape. Additional parameters are omitted for clarity but can be found in accompanying notebooks publicly available at \url{https://github.com/pavlin-policar/opentsne-paper}.

\section*{Discussion}

\subsection*{Uncovering Structure in High-Dimensional Data Sets}

Dimensionality reduction techniques implicitly assume that high-dimensional data lies on a lower-dimensional manifold, which can accurately be captured by a smaller number of dimensions. However, there is no evidence to suggest that any data set can accurately be described using only two dimensions and that any projection onto a two-dimensional plane will inevitably lead to loss of information. Therefore, it is beneficial to examine multiple embeddings, each of which provides a different perspective into the topology and other characteristics of the data. For instance, in Fig.~\ref{fig:tasic}, we show four embeddings of the same gene-expression data from the mouse brain~\cite{tasic2018shared}.
\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{tasic2018}
  \caption{\label{fig:tasic}We use openTSNE to create four different visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data set containing 21,874 single-cells originating from the mouse neocortex. Cluster annotations and colors are taken from the original publication. Warm colors correspond to excitatory neurons, cool colors correspond to inhibitory neurons, and gray/brown colors correspond to non-neuronal cells. The standard t-SNE \textbf{(a)} emphasizes local structure while increasing perplexity \textbf{(b)} results in a more meaningful layout of the clusters. We can also combine the two perplexities by using a multiscale kernel affinity model \textbf{(c)} and obtain a good trade-off between global and local structure. Alternatively, we can inspect more fine-grained structure and reveal smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}
Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters, which uncovers numerous clusters. Warm colors correspond to excitatory neurons, while cool colors represent inhibitory neurons. While different clusters of excitatory and inhibitory neurons appear close to one another, all clusters appear somewhat equidistant from their neighboring clusters, and the overall relations between them are not obvious. On the other hand, Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points, resulting in a more globally consistent layout where relations between clusters become more apparent. In this embedding, it is evident from the increased white-space between groups that there is one large class of excitatory neurons, and two related classes of inhibitory neurons. Unfortunately, focusing on preserving large neighborhoods often results in smaller clusters being absorbed into larger ones. Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels which aim to preserve both the global organization of clusters as well as prevent smaller clusters from being absorbed into larger ones. On the other hand, Fig.~\ref{fig:tasic}.d shows use the same parameter values as Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. This reveals that many of the clusters are composed of other subgroups.

When faced with a data set containing millions of data points, standard t-SNE embeddings often become unwieldy, with blurred cluster boundaries, large clusters absorbing smaller ones, and relationships between clusters becoming increasingly difficult to interpret. Fig.~\ref{fig:cao}.a contains expression profiles of over two million single cells captured at different time points in mouse development. The embedding indicates numerous clusters with some transition between time points as indicated by the color-coding, but this is largely difficult to interpret. Kobak \& Berens observed that increasing attractive forces between similar data points controlled via the \textit{exaggeration} parameter leads to more compact clusters, and subsequently more informative visualizations~\cite{kobak2019art}. For instance, \ref{fig:cao}.b doubles the default exaggeration  which begins to reveal the overall structure of the data. Doubling the exaggeration again in \ref{fig:cao}.c, we can now easily observe that the data is comprised of two main groups of cells and eight somewhat smaller clusters. The visualization also reveals a small number of very small clusters, possibly corresponding to rare cell types.

@@

Exaggeration can be used to highlight transitions between cell-states in developmental studies. Standard t-SNE often produces embeddings with clearly-defined, discrete clusters. We have shown that we can adjust the level of granularity and resolution of the clusters with several parameters. However, discrete clusters are often undesired in developmental studies where we assume cells follow a continuous path between cell states. To this end, other embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2} are often used, as they tend to better capture the continuity between cell states. Recently, Kobak \textit{et al.}~\cite{TODO} have shown that embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$ produce embeddings which are markedly similar to UMAP and ForceAtlas2, respectively. For example, in Fig.~\ref{fig:cao}.a, the developmental path between different time points is difficult to observe due to a large number of sprawled out clusters in the embedding. On the other hand, the developmental paths are easy to follow between cells in Fig.~\ref{fig:cao}.c, where we have increased the exaggeration factor to $4$.

Unlike other popular dimensionality reduction methods such as principal component analysis or autoencoders, t-SNE is a non-parametric method and does not define an explicit mapping to the embedding space. Therefore embeddings of new data points need to be found through the use of optimization techniques~\cite{policar2019embedding}. \opentsne\ is currently the only publicly available library allowing users to add new samples to existing embeddings.

Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a previously annotated single-cell data set and embed data from a separate experiment into the reference landscape. The reference data taken from Macosko \textit{et al.}~\cite{macosko2015highly} contains gene-expression profiles from mouse retinal cells. Embedding samples from a similar experiment focusing only on bipolar retinal cells by Shekhar \textit{et al.}~\cite{shekhar2016comprehensive}, we can observe that the majority of new cells correctly map to the bipolar cell clusters in the reference.

Embedding single cells into existing reference atlases can also be useful for cell-type classification in cases where cell identities are not known in advance. For instance, in Fig.~\ref{fig:transform}, we construct a reference embedding using annotated data from Hochgerner \textit{et al.}~\cite{hochgerner2018conserved} containing gene-expression profiles of cells from the mouse brain. The authors provide annotations for each cell, and we can verify their accuracy by visualizing well-established gene markers' major cell types. We then embed cells from Harris \textit{et al.}~\cite{harris2018classes}, where annotations are provided only for neuronal cells. We quickly identify other non-neuronal cell types, including oligodendrocytes and astrocytes. We use marker genes to validate that new cells were correctly mapped into the reference landscape.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{transform_hochgerner}
  \caption{\label{fig:transform}openTSNE enables the user to embed new samples into an existing, reference embedding. We construct a t-SNE embedding for the data from Hochgerner \textit{et al.}~\cite{hochgerner2018conserved} containing 24,185 developing, single cells from the mouse hippocampus. The data contains different types of neurons, supporting glia, and other vascular cells (upper left). Cell annotations are taken from the original publications. The colors correspond to Fig.~\ref{fig:tasic}. We then embed new, hippocampal cells collected by Harris \textit{et al.}~\cite{harris2018classes} using the previous embedding as a reference. In their study, Harris \textit{et al.} collected 6,971 single-cells and focused on identifying different types of inhibitory neurons. However, almost half of the collected cells are not neurons and were left uncharacterized. Inspecting where these cells are embedded in the reference embedding (bottom left) reveals that in addition to inhibitory neurons, the data contains several supporting glial cells as well as a small population of endothelial cells. We can verify the accuracy of our approach by inspecting marker genes for the major cell types in the reference (top row) and embedded samples (bottom row).}
\end{figure*}

These two examples demonstrate how \opentsne\ can be used to quickly gain insight into newly-sequenced, single-cell data sets by utilizing already existing, publicly-available, reference atlases. This approach is extremely general and is not limited to single-cell gene-expression data, but can be applied to any tabular data sets from any domain of science.

\subsection*{Versatility}

Versatility is another of  \opentsne s core design principles. Kobak \& Berens recently introduced several recommendations and tricks to obtain better and more meaningful t-SNE visualizations~\cite{kobak2019art}. These include the use of multi-scale similarity kernels, perplexity annealing, and increasing exaggeration when working with larger data sets. \opentsne\ provides an extremely flexible API, where these improvements can be incorporated in just a few lines of code. Furthermore, \opentsne\ supports custom affinity models, e.g. multi-scale Gaussian mixtures or uniform kernels, allowing for quick experimentation. This also enables users to construct t-SNE embeddings on non-tabular data, e.g. relational data, as the only affinity-model requirement is some notion of similarity between data points. Finally, our comprehensive callback system can be utilized to monitor and adapt different stages of the optimization phase and has been used to construct visually appealing animations of the t-SNE optimization process.

\subsection*{Speed}

One of t-SNE's common criticisms is poor scalability to large data sets containing up to millions of data points~\cite{becht2019dimensionality}. These criticisms are largely based on two factors. Until quite recently, most popular implementations of t-SNE were based on the Barnes-Hut approximation scheme developed by van der Maaten in 2014~\cite{van2014accelerating} with asymptotic complexity $\mathcal{O}(N \log N)$. More importantly, this criticism was exacerbated further by the fact that the most widely-used implementation of t-SNE came from \textsf{scikit-learn}, which exhibits very poor runtime performance compared to its C++ counterpart -- \textsf{MulticoreTSNE}. The multi-threaded \textsf{MulticoreTSNE} implementation~\cite{Ulyanov2016} is perfectly capable of creating t-SNE embeddings of millions of data points in a matter of hours on widely-accessible, consumer-grade processors (Fig.~\ref{fig:benchmarks}). However, recently, Linderman \textit{et al.} developed a new approximation scheme -- FIt-SNE -- which further reduces the asymptotic time complexity to $\mathcal{O}(N)$, enabling the embedding of large data sets in a matter of minutes.

Fig.~\ref{fig:benchmarks} shows benchmarks of four Python t-SNE implementations, including those from \textsf{scikit-learn} (v0.23.1), \textsf{MulticoreTSNE} (v0.1), \textsf{FIt-SNE} (v1.1.0), and \opentsne\ (v0.4.3). We perform benchmarks on two processors, one representing usage on a personal computer and one on a high-performance computing platform. The Intel(R) Core i7 is commonly found in consumer-grade laptop computers, while Intel(R) Xeon(R) processors are found in high-performance computing machines. Benchmarks were run with the original t-SNE parameters for 1000 iterations, as some implementations do not allow their modification.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{benchmarks}
  \caption{\label{fig:benchmarks}We benchmark openTSNE (v0.4.3) against three popular open-source implementations including scikit-learn~\cite{pedregosa2011scikit} (v0.23.1), MulticoreTSNE~\cite{Ulyanov2016} (v0.1), and FIt-SNE~\cite{linderman2019fast} (v1.1.0). Benchmarks were run on a consumer-grade Intel Core i7-7700HQ processor found in laptop computers, and on a server-grade Intel Xeon E5-2650. To generate benchmark data sets of different sizes, we subsampled data from the 10X Genomics 1.3 million mouse brain data set five times, resulting in five different data sets for each size. In total, each implementation was run on 30 different data sets. Notice that openTSNE scales similarly to FIt-SNE, as they both use the same interpolation-based approximation scheme, while scikit-learn and MulticoreTSNE utilize the Barnes-Hut approximation.}
\end{figure*}

Fig.~\ref{fig:benchmarks} demonstrates that both \textsf{FIt-SNE} and \opentsne\ exhibit better runtime performance than their Barnes-Hut counterparts -- \textsf{scikit-learn} and \textsf{MulticoreTSNE}. Notice that even though \textsf{FIt-SNE} and \opentsne\ both implement the same approximation scheme, \opentsne\ tends to be slightly slower. \opentsne\ is written in Python, which inevitably incurs some runtime overhead compared to its pure C++ counterpart. However, Fig.~\ref{fig:benchmarks} demonstrates the runtime of \opentsne\ is comparable to that of \textsf{FIt-SNE} and much faster than its Barnes-Hut competitors. Additionally, modern computer processors generally contain multiple cores, allowing us to make use of multi-threading. This further reduces the gap between \opentsne\ and \textsf{FIt-SNE}. On the server-grade processor, \opentsne\ is slightly faster than its pure C++ counterpart. We speculate that this is could be because \opentsne\ uses \textsf{numpy} for most linear algebra operations, which may make better use of the Intel Math Kernel Library (MKL), which are aggressively optimized on Intel(R) Xeon(R) processors.

Additionally, \opentsne\ provides a flexible API, which allows splitting up the embedding-construction process into several part, caching slow operations. This enables users to quickly experiment with different parameter settings and iterate on their final visualizations.

Widespread adoption of novel techniques is almost universally accompanied by easy-to-use and easy-to-install software. While the t-SNE implementation from \textsf{scikit-learn} fits this requirement, its implementation is unfortunately very slow, making it unsuitable for its application to large data sets. Other C++ implementations such as \textsf{MulticoreTSNE} and \textsf{FIt-SNE} exhibit better scaling in larger data sets, but do not provide precompiled binaries, requiring users to compile the software themselves. This is problematic from a usability standpoint, as many users often lack the knowledge to do so. This is especially problematic with Windows users, where the operating system does not include a C++ compiler by default, and correctly configuring the command line can be an arduous process.

\opentsne\ is a Python package that focuses on accessibility. We provide precompiled binaries for all major Python versions on all major platforms, making the installation process as seamless as possible. \opentsne\ can be installed through the Python Package Index (PyPI) or through conda on the conda-forge channel. \opentsne\ is designed to be familiar to Python users and follows the \textsf{scikit-learn} API~\cite{sklearn_api}, which is well established in the Python data science ecosystem. \opentsne\ implements multi-threaded versions of both the Barnes-Hut and FIt-SNE approximation schemes, allowing it to be used on data sets containing millions of data points. While the Python virtual machine introduces some performance overhead, the runtime is comparable to that of its pure C++ counterpart -- \textsf{FIt-SNE}. Finally, \opentsne\ is extensible. Its modular design enables researchers to quickly experiment with different settings and easily incorporate their own components into the software.

@@

This section is meant to familiarize the reader with the t-SNE algorithm and its recent extensions. This brief review is by no means comprehensive, but is self contained and introduces the mathematics upon which the \opentsne\ library is based.

\subsection*{Preliminaries}
This section is meant to familiarize the reader with the t-SNE algorithm and its recent extensions. This brief review is by no means comprehensive, but is self-contained and introduces the mathematics upon which the \opentsne\ library is based.

\subsection*{Preliminaries}

t-distributed stochastic neighbor embedding (t-SNE) is a non-linear dimensionality reduction method, which aims to find a low-dimensional embedding, where local neighborhoods are preserved. More formally, given a multi-dimensional data set $\mathbf{X} = \left \{ \mathbf{x}_1,
\mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in \mathbb{R}^D$ where $N$ is the
number of data points in the data set, t-SNE aims to find a low
dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1, \mathbf{y}_2, \dots,
\mathbf{y}_N \right \} \in \mathbb{R}^d$ where $d \ll D$, such that if points
$\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the high-dimensional space,
their corresponding embeddings $\mathbf{y}_i$ and $\mathbf{y}_j$ are also
close. Since t-SNE is primarily used as a visualization tool, $d$ is typically
set to two. The similarity between two data points in the high-dimensional space is defined as:

@@

\noindent where $\mathcal{D}$ is some distance measure. This is then symmetrized to

@@

\noindent Different bandwidths $\sigma_i$ enable t-SNE to adapt to the varying
density of the data in the multi-dimensional space. Perplexity is sometimes considered the continuous analog to the $k$ nearest neighbors to which distances will be preserved.

The similarity between points $\mathbf{y}_i$ and $\mathbf{y}_j$ in the
embedding space is defined using the $t$-distribution with a single degree of
freedom. This is also known as the Cauchy kernel:

@@

Finally, the Kullback-Leibler (KL) divergence is used as a measure of agreement between distributions $\mathbf{P}$ and $\mathbf{Q}$

@@

\noindent The t-SNE objective is to find embeddings $\mathbf{Y}$ such that the KL divergence is minimized.

The corresponding gradient takes the form
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j || ^2 \right )^{-1}.
\label{eq:tsne_gradient}
\end{equation}

Optimization is performed using batch gradient descent with the delta-bar-delta update rule~\cite{jacobs1988increased}. Originally, t-SNE was run for 1000 iterations consisting of two phases. In the first \textit{early exaggeration} phase, the attractive forces between data points is increased by a factor $\rho=12$ so that points can more easily move throughout the embedding to find their respective neighbors. The remaining 750 iterations are run with $\rho=1$, which reverts the attractive forces to their original values and produces the final embedding.

Belkina \textit{et al.} later found that faster convergence can be achieved by increasing the learning rate to $\eta=N/12$~\cite{belkina2019automated}. The number of iterations can then safely be lowered to 750, leading to faster runtime. This convention has been adopted by most modern t-SNE implementations.

@@

Van der Maaten observed that since $p_{ij}$ is calculated in the high-dimensional space using a Gaussian kernel and that the bandwidth is selected such that only a fixed number of closest neighbors (determined by the perplexity parameter) will contribute to the attractive forces of a given point due to the exponential decay of the Gaussian kernel~\cite{van2014accelerating}. Therefore it is sufficient to calculate the attractive forces only over a relatively small number of nearest neighbors instead of all $N$ points. By utilizing tree-based nearest-neighbor search methods, the time complexity is thus reduced to $\mathcal{O}(N \log N)$. Linderman \textit{et al.} further realized that, qualitatively, embeddings are visually virtually indistinguishable when using only \textit{approximate} nearest neighbors, further reducing time complexity to $\mathcal{O}(N)$~\cite{linderman2019fast}.

@@

@@

In standard t-SNE $\alpha=1$ so this simplifies to the Cauchy kernel from Eqn.~(\ref{eq:cauchy_kernel}).
